Diagnostic Potential \& the Interactive Dynamics of the Colorectal Cancer Virome

### 2018-09-30 doc/oman.tex

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{Helvetica}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={The Diagnostic Potential \& Interactive Dynamics of the Colorectal Cancer Virome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{parskip}
\setlength{\parskip}{7.5pt}
\exhyphenpenalty=10000
\hyphenpenalty=10000

@@

  \title{The Diagnostic Potential \& Interactive Dynamics of the Colorectal
Cancer Virome}
    \usepackage{authblk}
    \font\myfont=Helvetica at 12pt
    \font\nextfont=Helvetica at 10pt
                        \author[1]{\myfont Geoffrey D Hannigan}
                    \author[2]{\myfont Melissa B Duhaime}
                    \author[3]{\myfont Mack T Ruffin IV}
                    \author[1]{\myfont Charlie C Koumpouras}
                    \author[1,*]{\myfont Patrick D Schloss}
                            \affil[1]{\nextfont Department of Microbiology \& Immunology, University of Michigan, Ann
Arbor, Michigan, 48109}
                    \affil[2]{\nextfont Department of Ecology and Evolutionary Biology, University of Michigan,
Ann Arbor, Michigan, 48109}
                    \affil[3]{\nextfont Department of Family and Community Medicine, Pennsylvania State
University Hershey Medical Center, Hershey, Pennsylvania, 17033}
                    \affil[*]{\nextfont To whom correspondence may be addressed.}
            \date{}

@@

Human viruses (those that infect human cells) have been associated with
many cancers, largely due to their mutagenic and functionally
manipulative abilities. Despite this, cancer microbiome studies have
almost exclusively focused on bacteria instead of viruses. We began
evaluating the cancer virome by focusing on colorectal cancer, a primary
cause of morbidity and mortality throughout the world, and a cancer
linked to altered colonic bacterial community compositions but with an
unknown association with the gut virome. We used 16S rRNA gene, whole
shotgun metagenomic, and purified virus metagenomic sequencing of stool
to evaluate the differences in human colorectal cancer virus and
bacterial community composition. Through random forest modeling we
identified differences in the healthy and colorectal cancer virome. The
cancer-associated virome consisted primarily of temperate bacteriophages
that were also predicted to be bacteria-virus community network hubs.
These results provide foundational evidence that bacteriophage
communities are associated with colorectal cancer and potentially impact
cancer progression by altering the bacterial host communities.

@@

Colorectal cancer is a leading cause of cancer-related death in the
United States and worldwide. Its risk and severity have been linked to
colonic bacterial community composition. Although human-specific viruses
have been linked to other cancers and diseases, little is known about
colorectal cancer virus communities. We addressed this knowledge gap by
identifying differences in colonic virus communities in the stool of
colorectal cancer patients and how they compared to bacterial community
differences. The results suggested an indirect role for the virome in
impacting colorectal cancer by modulating their associated bacterial
community. These findings both support a biological role for viruses in
colorectal cancer and provide a new understanding of basic colorectal
cancer etiology.

@@

The human gut virome is the community of all viruses found in the gut,
including bacteriophages (viruses that only infect bacteria), eukaryotic
viruses (viruses that only infect eukaryotic cells), and human-specific
viruses (viruses that only infect human cells). Due to their mutagenic
abilities and propensity for functional manipulation, human viruses are
strongly associated with, and in many cases cause, cancer (1--4).
Because bacteriophages are crucial for bacterial community stability and
composition (5--7), and members of those bacterial communities have been
implicated as oncogenic agents (8--11), bacteriophages have the
potential to indirectly impact cancer as well. The gut virome therefore
has a potential to be associated with, and potentially impact, human
cancer. Altered human virome composition and diversity have already been
identified in diseases including periodontal disease (12), HIV (13),
cystic fibrosis (14), antibiotic exposure (15, 16), urinary tract
infections (17), and inflammatory bowel disease (18). The strong
association of bacterial communities with colorectal cancer, the
previous identification of human-specific viruses that cause cancer, and
the precedent for the virome to impact other human diseases suggest that
colorectal cancer may be associated with altered virus communities.

@@

Here we begin to address the knowledge gap of whether virus community
composition is altered in colorectal cancer and, if it is, how those
differences might impact cancer progression and severity. We also aimed
to evaluate the virome's potential for use as a diagnostic biomarker.
The implications of this study are threefold. \emph{First}, this work
supports a biological role for the virome in colorectal cancer
development and suggests that more than the bacterial members of the
associated microbial communities are involved in the process.
\emph{Second}, we present a supplementary virus-based approach for
classification modeling of colorectal cancer using stool samples.
\emph{Third}, we provide initial support for the importance of studying
the virome as a component of the microbiome ecological network,
especially in cancer.

@@

Our study cohort consisted of stool samples collected from 90 human
subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30
of whom had carcinomas \textbf{(Figure \ref{sampleproc})}. Half of each
stool sample was used to sequence the bacterial communities using both
16S rRNA gene and shotgun sequencing techniques. The 16S rRNA gene
sequencing was performed for a previous study, and the sequences were
re-analyzed using contemporary methods (8). The other half of each stool
sample was purified for virus like particles (VLPs) before genomic DNA
extraction and shotgun metagenomic sequencing. In the VLP purification,
cells were disrupted and extracellular DNA degraded \textbf{(Figure
\ref{sampleproc})} to allow the exclusive analysis of viral DNA within
virus capsids. In this manner, the \emph{extracellular virome} of
encapsulated viruses was targeted.

@@

Microbiome and disease associations are often described as being of an
altered diversity (i.e. ``dysbiotic''). Therefore, we first evaluated
the influence of colorectal cancer on virome OVU diversity. We evaluated
differences in communities between disease states using the Shannon
diversity, richness, and Bray-Curtis metrics. We observed no significant
alterations in either Shannon diversity or richness in the diseased
states as compared to the healthy state \textbf{(Figure \ref{betaogu}
C-D)}. There was no statistically significant clustering of the disease
groups (ANOSIM p-value = 0.6, \textbf{Figure \ref{betaogu}}). Notably,
there was a significant difference between the few blank controls that
remained after rarefying the data and the other study groups (ANOSIM
p-value \textless{} 0.001, \textbf{Figure \ref{betaogunegative})},
further supporting the quality of the sample set. In summary, standard
alpha and beta diversity metrics were insufficient for capturing virus
community differences between disease states \textbf{(Figure
\ref{betaogu})}. This is consistent with what has been observed when the
same metrics were applied to 16S rRNA gene sequences and metagenomic
samples (8, 23, 24) and points to the need for alternate approaches to
detect the impact of colorectal cancer disease state on these community
structures.

@@

As opposed to the diversity metrics discussed above, OTU-based relative
abundance profiles generated from 16S rRNA gene sequences are effective
for classifying stool samples as originating from individuals with
healthy, adenomatous, or cancerous colons (8, 23). By using
classification models instead of attempting to identify single
differentially abundant OTUs, these and other studies have been
successful in capturing complex community relationships in which
differences in taxonomic relative abundance are considered in the
context of other taxa. The exceptional performance of bacteria in these
classification models supports a role for bacterial functionality in
colorectal cancer. We built off of these findings by evaluating the
ability of virus community signatures to classify stool samples and
compared their performance to models built using bacterial community
signatures.

To identify the altered virus communities associated with colorectal
cancer, we built and tested random forest models for classifying stool
samples as belonging to individuals with either cancerous or healthy
colons. We confirmed that our bacterial 16S rRNA gene model replicated
the performance of the original report which used logit models instead
of random forest models \textbf{(Figure \ref{predmodel} A)} (8). We then
compared the bacterial OTU model to a model built using OVU relative
abundances. The viral model performed as well as the bacterial model
(corrected p-value = 0.4), with the viral and bacterial models achieving
mean area under the curve (AUC) values of 0.792 and 0.809, respectively
\textbf{(Figure \ref{predmodel} A - B)}. To evaluate the ability of both
bacterial and viral biomarkers to classify samples, we built a combined
model that used both bacterial and viral community data. The combined
model did not yield a statistically significant performance improvement
beyond the viral (corrected p-value = 0.4) and bacterial (corrected
p-value = 0.08) models, yielding an AUC of 0.768 \textbf{(Figure
\ref{predmodel} A - B)}.

We compared viral metagenomic methods to bacterial metagenomic methods
by building a viral model and a model built using OGU relative abundance
profiles from bacterial metagenomic shotgun sequencing data. This
bacterial model performed worse than the other models (mean AUC = 0.474)
\textbf{(Figure \ref{predmodel} A - B)}. To determine the cause of the
discrepancy between the two bacterial sequencing methods, we attempted
to compare the approaches at a common sequencing depth. This revealed
that the bacterial 16S rRNA gene model was strongly driven by sparse and
low abundance OTUs \textbf{(Figure \ref{16scompare})}. Removal of OTUs
with a median abundance of zero resulted in the removal of six OTUs, and
a loss of model performance down to what was observed in the
metagenome-based model \textbf{(Figure \ref{16scompare} A)}. The
majority of these OTUs had a relative abundance lower than 1\% across
the samples \textbf{(Figure \ref{16scompare} B)}. Although the features
in the viral model also were of low abundance \textbf{(Figure
\ref{threewaymodel} F)}, the coverage was sufficient for high model
performance, likely because viral genomes are orders of magnitude
smaller than bacterial genomes.

The association between the bacterial and viral communities and
colorectal cancer was driven by a few important microbes.
\emph{Fusobacterium} was the primary driver of the bacterial association
with colorectal cancer, which is consistent with its previously
described oncogenic potential \textbf{(Figure \ref{predmodel} C)}(22).
The virome signature also was driven by a few OVUs, suggesting a role
for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
It is also important to note that while these viruses were driving the
signature, the magnitude of their importance and the significance of
those values was noticibly less than the bacterial 16S signature,
suggesting that unlike what is observed in the bacteria, there are many
viruses that are associated with the cancerous state. The identified
viruses were bacteriophages, belonging to \emph{Siphoviridae},
\emph{Myoviridae}, and phage taxa that could not be confidently
identified beyond their broad phage identification (i.e.
``unclassified''). Many of the important viruses, that were confirmed to
not have genomic similarity to known bacterial genomes, were
unidentifiable (denoted ``unknown''). This is common in viromes across
habitats; studies have reported as much as 95\% of virus sequences
belonging to unknown genomic units (14, 28--30). When the bacterial and
viral community signatures were combined, both bacterial and viral
organisms drove the community association with cancer \textbf{(Figure
\ref{predmodel} E)}.

@@

Because previous work has identified shifts in which bacteria were most
important at different stages of colorectal cancer (8, 20, 22), we
explored whether shifts in the relative influence of phages could be
detected between healthy, adenomatous, and cancerous colons. We
evaluated community shifts between the disease stage transitions
(healthy to adenomatous and adenomatous to cancerous) by building random
forest models to compare only the diagnosis groups around the
transitions. While bacterial OTU models performed equally well for all
disease class comparisons, the virome model performances differed
\textbf{(Figure \ref{transitionmodels} A-B)}. Like bacteria
\textbf{(Figure \ref{transitionmodels} F-H)}, different virome members
were important between the healthy to adenomatous and adenomatous to
cancerous stages \textbf{(Figure \ref{transitionmodels} C-E)}.

After evaluating our ability to classify samples between two disease
states, we performed a three-class random forest model including all
disease states. The 16S rRNA gene model yielded a mean AUC of 0.765 and
outperformed the viral community model, which yielded a mean AUC of
0.658 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
A-C}). The microbes important for the healthy versus cancer and healthy
versus adenoma models were also important for the three-class model
\textbf{(Figure \ref{threewaymodel} D-E)}. The most important bacterium
in the two and three class models was the same \emph{Fusobacterium} (OTU
4) \textbf{(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)}.
The viruses most important to the three-class model were identified as
bacteriophages \textbf{(Figure \ref{predmodel} D, Figure
\ref{threewaymodel} E)}, but not all important OVUs were of increased
abundance in the diseased state \textbf{(Figure \ref{threewaymodel} F)}.

@@

Differences in the colorectal cancer virome could have been driven by
eukaryotic (human) viruses or by bacteriophages. To better understand
the types of viruses that were important for colorectal cancer, we
identified the virome OVUs as being similar to either eukaryotic viruses
or bacteriophages. The most important viruses to the classification
model were identified as bacteriophages (\textbf{Figure
\ref{threewaymodel})}. Overall, we were able to identify 78.8\% of the
OVUs as known viruses, and 93.8\% of those viral OVUs aligned to
bacteriophage reference genomes. It is important to note that this could
have been influenced by our methodological biases against enveloped
viruses (more common of eukaryotic viruses than bacteriophage), due to
chloroform and DNase treatment for purification.

We evaluated whether the phages in the community were primarily lytic
(i.e.~obligately lyse their hosts after replication) or temperate
(i.e.~able to integrate into their host's genome to form a lysogen, and
subsequently transition to a lytic mode). We accomplished this by
identifying three markers for temperate phages in the OVU representative
sequences: 1) presence of phage integrase genes, 2) presence of known
prophage genes, according the the ACLAME (A CLAssification of Mobile
genetic Elements) database, and 3) nucleotide similarity to regions of
bacterial genomes (29, 31, 32). We found that the majority of the phages
were temperate and that the overall fraction of temperate phages
remained consistent throughout the healthy, adenomatous, and cancerous
stages \textbf{(Figure \ref{replicationstyles})}. These findings were
consistent with previous reports suggesting the gut virome is primarily
composed of temperate phages (13, 18, 31, 33).

@@

Because the link between colorectal cancer and the virome was driven by
bacteriophages (as opposed to non-bacterial viruses), we tested a
potential hypothesis that the virome signal was a mere reflection of the
bacterial signal, and thus highly correlated with the bacterial signal.
If this hypothesis were true, we would expect a correlation between the
relative abundances of influential bacterial OTUs and virome OVUs.
Instead, we observed a strikingly low correlation between bacterial and
viral relative abundances \textbf{(Figure \ref{correlations} A,C)}.
Overall, there was an absence of correlation between the most
influential OVUs and bacterial OTUs \textbf{(Figure \ref{correlations}
B)}. This evidence supported our null hypothesis that the influential
viral OVUs were not primarily reflections of influential bacteria.

Given these findings, we posited that the most influential phages were
acting by infecting a wide range of bacteria in the overall community,
instead of just the influential bacteria. In other words, we
hypothesized that the influential bacteriophages were community hubs
(i.e.~central members) within the bacteria and phage interactive
network. We investigated the potential host ranges of all phage OVUs
using a previously developed random forest model that relies on sequence
features to predict which phages infected which bacteria in the
community \textbf{(Figure \ref{network} A)} (34). The predicted
interactions were then used to identify phage community hubs. We
calculated the alpha centrality (i.e.~measure of importance in the
ecological network) of each phage OVU's connection to the rest of the
network. The phages with high centrality values were defined as
community hubs. Next, the centrality of each OVU was compared to its
importance in the colorectal cancer classification model. Phage OVU
centrality was significantly and positively correlated with importance
to the disease model (p-value = 0.004, R = 0.176), suggesting that
phages that were important in driving colorectal cancer also were more
likely to be community hubs \textbf{(Figure \ref{network} B)}. Together
these findings supported our hypothesis that influential phages were
hubs within their microbial communities and had broad host ranges.

@@

https://github.com/SchlossLab/Hannigan\_CRCVirome\_mBio\_2018

@@

The authors thank the Schloss lab members for their underlying
contributions, and the Great Lakes-New England Early Detection Research
Network for providing the fecal samples that were used in this study.
The authors also thank the participants in the study cohort.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{Helvetica}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Diagnostic Potential \& the Interactive Dynamics of the Colorectal Cancer Virome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{parskip}
\setlength{\parskip}{7.5pt}
\exhyphenpenalty=10000
\hyphenpenalty=10000

@@

  \title{Diagnostic Potential \& the Interactive Dynamics of the Colorectal
Cancer Virome}
    \usepackage{authblk}
    \font\myfont=Helvetica at 12pt
    \font\nextfont=Helvetica at 10pt
                        \author[1]{\myfont Geoffrey D Hannigan}
                    \author[2]{\myfont Melissa B Duhaime}
                    \author[3]{\myfont Mack T Ruffin IV}
                    \author[1]{\myfont Charlie C Koumpouras}
                    \author[1,*]{\myfont Patrick D Schloss}
                            \affil[1]{\nextfont Department of Microbiology \& Immunology, University of Michigan, Ann
Arbor, Michigan, 48109}
                    \affil[2]{\nextfont Department of Ecology and Evolutionary Biology, University of Michigan,
Ann Arbor, Michigan, 48109}
                    \affil[3]{\nextfont Department of Family and Community Medicine, Pennsylvania State
University Hershey Medical Center, Hershey, Pennsylvania, 17033}
                    \affil[*]{\nextfont To whom correspondence may be addressed.}
            \date{}

@@

Viruses are associated with many human cancers, largely due to their
mutagenic and functionally manipulative abilities. Despite this, cancer
microbiome studies have almost exclusively focused on bacteria instead
of viruses. We began evaluating the cancer virome by focusing on
colorectal cancer, a primary cause of morbidity and mortality throughout
the world, and a cancer linked to altered colonic bacterial community
compositions while the virome role remains unknown. We used 16S rRNA
gene, whole shotgun metagenomic, and purified virus metagenomic
sequencing of stool to evaluate the differences in human colorectal
cancer virus and bacterial community composition. Through random forest
modeling we identified differences in the healthy and colorectal cancer
virome. The cancer-associated virome consisted primarily of temperate
bacteriophages that were also predicted to be bacteria-virus community
network hubs. These results provide foundational evidence that
bacteriophage communities are associated with colorectal cancer and
likely impact cancer progression by altering the bacterial host
communities.

@@

Colorectal cancer is a leading cause of cancer-related death in the
United States and worldwide. Its risk and severity have been linked to
colonic bacterial community composition. Although viruses have been
linked to other cancers and diseases, little is known about colorectal
cancer virus communities. We addressed this knowledge gap by identifying
differences in colonic virus communities in the stool of colorectal
cancer patients and how they compared to bacterial community
differences. The results suggested an indirect role for the virome in
impacting colorectal cancer by modulating their associated bacterial
community. These findings both support a biological role for viruses in
colorectal cancer and provide a new understanding of basic colorectal
cancer etiology.

@@

Due to their mutagenic abilities and propensity for functional
manipulation, human viruses are strongly associated with, and in many
cases cause, cancer (1--4). Because bacteriophages (i.e.~viruses that
specifically infect bacteria) are crucial for bacterial community
stability and composition (5--7) and have been implicated as oncogenic
agents (8--11), bacteriophages have the potential to indirectly impact
cancer. The gut virome (i.e.~the virus community of the gut) therefore
has the potential to impact health and disease. Altered human virome
composition and diversity have been identified in diseases including
periodontal disease (12), HIV (13), cystic fibrosis (14), antibiotic
exposure (15, 16), urinary tract infections (17), and inflammatory bowel
disease (18). The strong association of bacterial communities with
colorectal cancer and the precedent for the virome to impact other human
diseases suggest that colorectal cancer may be associated with altered
virus communities.

@@

Here we address the knowledge gap of whether virus community composition
is altered in colorectal cancer and, if it is, how those differences
might impact cancer progression and severity. We also aimed to evaluate
the virome's potential for use as a diagnostic biomarker. The
implications of this study are threefold. \emph{First}, this work
supports a biological role for the virome in colorectal cancer
development and suggests that more than the bacterial members of the
associated microbial communities are involved in the process.
\emph{Second}, we present a supplementary, or even alternative,
virus-based approach for classification modeling of colorectal cancer
using stool samples. \emph{Third}, we provide initial support for the
importance of studying the virome as a component of the microbiome
ecological network, especially in cancer.

@@

Our study cohort consisted of 90 human subjects, 30 of whom had healthy
colons, 30 of whom had adenomas, and 30 of whom had carcinomas
\textbf{(Figure \ref{sampleproc})}. Half of each stool sample was used
to sequence the bacterial communities using both 16S rRNA gene and
shotgun sequencing techniques. The 16S rRNA gene sequencing was
performed for a previous study, and the sequences were re-analyzed using
contemporary methods (8). The other half of each stool sample was
purified for virus like particles (VLPs) before genomic DNA extraction
and shotgun metagenomic sequencing. In the VLP purification, cells were
disrupted and extracellular DNA degraded \textbf{(Figure
\ref{sampleproc})} to allow the exclusive analysis of viral DNA within
virus capsids. In this manner, the \emph{extracellular virome} of
encapsulated viruses was targeted.

@@

Microbiome and disease associations are often described as being of an
altered diversity (i.e., ``dysbiotic''). Therefore, we first evaluated
the influence of colorectal cancer on virome OVU diversity. We evaluated
differences in communities between disease states using the Shannon
diversity, richness, and Bray-Curtis metrics. We observed no significant
alterations in either Shannon diversity or richness in the diseased
states as compared to the healthy state \textbf{(Figure \ref{betaogu}
C-D)}. There was no statistically significant clustering of the disease
groups (ANOSIM p-value = 0.4, \textbf{Figure \ref{betaogu}}). Notably,
there was a significant difference between the few blank controls that
remained after rarefying the data and the other study groups (ANOSIM
p-value \textless{} 0.001, \textbf{Figure \ref{betaogunegative})},
further supporting the quality of the sample set. In summary, standard
alpha and beta diversity metrics were insufficient for capturing virus
community differences between disease states \textbf{(Figure
\ref{betaogu})}. This is consistent with what has been observed when the
same metrics were applied to 16S rRNA gene sequences and metagenomic
samples (8, 23, 24) and points to the need for alternate approaches to
detect the impact of colorectal cancer disease state on these
communities.

@@

As opposed to the diversity metrics discussed above, OTU-based relative
abundance profiles generated from 16S rRNA gene sequences are effective
for classifying stool samples as originating from individuals with
healthy, adenomatous, or cancerous colons (8, 23). The exceptional
performance of bacteria in these classification models supports a role
for bacteria in colorectal cancer. We built off of these findings by
evaluating the ability of virus community signatures to classify stool
samples and compared their performance to models built using bacterial
community signatures.

To identify the altered virus communities associated with colorectal
cancer, we built and tested random forest models for classifying stool
samples as belonging to individuals with either cancerous or healthy
colons. We confirmed that our bacterial 16S rRNA gene model replicated
the performance of the original report which used logit models instead
of random forest models \textbf{(Figure \ref{predmodel} A)} (8). We then
compared the bacterial OTU model to a model built using OVU relative
abundances. The viral model performed as well as the bacterial model
(corrected p-value = 0.7), with the viral and bacterial models achieving
mean area under the curve (AUC) values of 0.767 and 0.772, respectively
\textbf{(Figure \ref{predmodel} A - B)}. To evaluate the ability of both
bacterial and viral biomarkers to classify samples, we built a combined
model that used both bacterial and viral community data. The combined
model did not yield a statistically significant performance improvement
beyond the viral (corrected p-value = 0.1) and bacterial (corrected
p-value = 0.2) models, yielding an AUC of 0.804 \textbf{(Figure
\ref{predmodel} A - B)}.

To determine the advantage of viral metagenomic methods over bacterial
metagenomic methods, we compared the viral model to a model built using
OGU relative abundance profiles from bacterial metagenomic shotgun
sequencing data. This model performed worse than the other models (mean
AUC = 0.467) \textbf{(Figure \ref{predmodel} A - B)}. Because the
coverage provided by the metagenomic sequencing was not as deep as the
equivalent 16S rRNA gene sequencing, we attempted to compare the
approaches at a common sequencing depth. This investigation revealed
that the bacterial 16S rRNA gene model was strongly driven by sparse and
low abundance OTUs \textbf{(Figure \ref{16scompare})}. Removal of OTUs
with a median abundance of zero resulted in the removal of six OTUs, and
a loss of model performance down to what was observed in the
metagenome-based model \textbf{(Figure \ref{16scompare} A)}. The
majority of these OTUs had a relative abundance lower than 1\% across
the samples \textbf{(Figure \ref{16scompare} B)}. Although the features
in the viral model also were of low abundance \textbf{(Figure
\ref{threewaymodel} F)}, the coverage was sufficient for high model
performance, likely because viral genomes are orders of magnitude
smaller than bacterial genomes. Thus, the targeted 16S rRNA gene
sequencing approach, which represented only a fraction of the bacterial
metagenomic sequencing depth, was more effective for detecting
colorectal cancer in stool samples. Despite the recent loss of
enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic
techniques, 16S rRNA gene sequencing is still a superior methodological
approach for some important applications.

The association between the bacterial and viral communities and
colorectal cancer was driven by a few important microbes.
\emph{Fusobacterium} was the primary driver of the bacterial association
with colorectal cancer, which is consistent with its previously
described oncogenic potential \textbf{(Figure \ref{predmodel} C)}(22).
The virome signature also was driven by a few OVUs, suggesting a role
for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
The identified viruses were bacteriophages, belonging to
\emph{Siphoviridae}, \emph{Myoviridae}, and ``unclassified'' phage taxa.
Many of the important viruses were unidentifiable (denoted ``unknown'').
This is common in viromes across habitats; studies have reported as much
as 95\% of virus sequences belonging to unknown genomic units (14,
28--30). When the bacterial and viral community signatures were
combined, both bacterial and viral organisms drove the community
association with cancer \textbf{(Figure \ref{predmodel} E)}.

@@

Because previous work has identified shifts in which bacteria were most
important at different stages of colorectal cancer (8, 20, 22), we
explored whether shifts in the relative influence of specific phages
could be detected between healthy, adenomatous, and cancerous colons. We
evaluated community shifts between the two disease stage transitions
(healthy to adenomatous and adenomatous to cancerous) by building random
forest models to compare only the diagnosis groups around the
transitions. While bacterial OTU models performed equally well for all
disease class comparisons, the virome model performances differed
\textbf{(Figure \ref{transitionmodels} A-B)}. Like bacteria
\textbf{(Figure \ref{transitionmodels} F-H)}, different virome members
were important between the healthy to adenomatous and adenomatous to
cancerous stages \textbf{(Figure \ref{transitionmodels} C-E)}.

After evaluating our ability to classify samples between two disease
states, we performed a three-class random forest model including all
disease states. The 16S rRNA gene model yielded a mean AUC of 0.774 and
outperformed the viral community model, which yielded a mean AUC of
0.672 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
A-C}). The microbes important for the healthy versus cancer and healthy
versus adenoma models were also important for the three-class model
\textbf{(Figure \ref{threewaymodel} D-E)}. The most important bacterium
in the two and three class models was the same \emph{Fusobacterium} (OTU
4) \textbf{(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)}.
The viruses most important to the three-class model were identified as
bacteriophages \textbf{(Figure \ref{predmodel} D, Figure
\ref{threewaymodel} E)}, but not all important OVUs were of increased
abundance in the diseased state \textbf{(Figure \ref{threewaymodel} F)}.

@@

Differences in the colorectal cancer virome could have been driven
directly by eukaryotic viruses or indirectly by bacteriophages acting
through their bacterial hosts. To better understand the types of viruses
that were important for colorectal cancer, we identified the virome OVUs
as being similar to either eukaryotic viruses or bacteriophages. The
most important viruses to the classification model were identified as
bacteriophages (\textbf{Figure \ref{threewaymodel})}. Overall, we were
able to identify 78.8\% of the OVUs as known viruses, and 93.8\% of
those viral OVUs aligned to bacteriophage reference genomes. It is
important to note that this could have been influenced by our
methodological biases against enveloped viruses (more common of
eukaryotic viruses than bacteriophage), due to chloroform and DNase
treatment for purification.

We evaluated whether the phages in the community were primarily lytic
(i.e.~obligately lyse their hosts after replication) or temperate
(i.e.~able to integrate into their host's genome to form a lysogen, and
subsequently transition to a lytic mode). We accomplished this by
identifying three markers for temperate phages in the OVU representative
sequences: 1) presence of phage integrase genes, 2) presence of known
prophage genes, according the the ACLAME (A CLAssification of Mobile
genetic Elements) database, and 3) nucleotide similarity to regions of
bacterial genomes (29, 31, 32). We found that the majority of the phages
were temperate and that the overall fraction of temperate phages
remained consistent throughout the healthy, adenomatous, and cancerous
stages \textbf{(Figure \ref{replicationstyles} E)}. These findings were
consistent with previous reports suggesting the gut virome is primarily
composed of temperate phages (13, 18, 31, 33).

@@

Because the link between colorectal cancer and the virome was driven by
bacteriophages, we hypothesized that the influential phages were
primarily predators of the influential bacteria, and thus influenced
their relative abundance through predation. If this hypothesis were
true, we would expect a correlation between the relative abundances of
influential bacteria and phages. Instead, we observed a strikingly low
correlation between bacterial and phage relative abundances
\textbf{(Figure \ref{correlations} A,C)}. Overall, there was an absence
of correlation between the most influential OVUs and bacterial OTUs
\textbf{(Figure \ref{correlations} B)}. This evidence supported our null
hypothesis that the influential phages were not primarily predators of
influential bacteria.

Given these findings, we hypothesized that the most influential phages
were acting by infecting a wide range of bacteria in the overall
community, instead of just the influential bacteria. In other words, we
hypothesized that the influential bacteriophages were community hubs
(i.e.~central members) within the bacteria and phage interactive
network. We investigated the potential host ranges of all phage OVUs
using a previously developed random forest model that relies on sequence
features to predict which phages infected which bacteria in the
community \textbf{(Figure \ref{network} A)} ({\textbf{???}}). The
predicted interactions were then used to identify phage community hubs.
We calculated the alpha centrality (i.e.~measure of importance in the
ecological network) of each phage OVU's connection to the rest of the
network. The phages with high centrality values were defined as
community hubs. Next, the centrality of each OVU was compared to its
importance in the colorectal cancer classification model. Phage OVU
centrality was significantly and positively correlated with importance
to the disease model (p-value = 0.02, R = 0.14), suggesting that phages
important in driving colorectal cancer also were more likely to be
community hubs \textbf{(Figure \ref{network} B)}. Together these
findings supported our hypothesis that influential phages were hubs
within their microbial communities and had broad host ranges.

@@

https://github.com/SchlossLab/Hannigan\_CRCVirome\_mBio\_2017

@@

The authors thank the Schloss lab members for their underlying
contributions, and the Great Lakes-New England Early Detection Research
Network for providing the fecal samples that were used in this study.
